---
title: "Lab 4: Data Wrangling"
author: "Environmental Data Analytics | John Fay and Luana Lima | Developed by Kateri Salk"
date: "Spring 2022"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
1- Answer questions on M3/A3
2- Answer questions on M4
3- Practice wrangling datasets with dplyr functions

## Set up your session
```{r, message = FALSE }
getwd()
setwd("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022/Lessons")
```

Today we will work with a dataset from the [North Temperate Lakes Long-Term Ecological Research Station](https://lter.limnology.wisc.edu/about/overview). The NTL-LTER is located in the boreal zone in northern Wisconsin, USA. We will use the [chemical and physical limnology dataset](https://lter.limnology.wisc.edu/content/cascade-project-north-temperate-lakes-lter-core-data-physical-and-chemical-limnology-1984), running from 1984-2016. 

Opening discussion: why might we be interested in long-term observations of temperature, oxygen, and light in lakes?

> Add notes here: 

```{r, message = FALSE}
install.packages(tidyverse)
library(tidyverse)
install.packages(lubridate)
library(lubridate)
NTL.phys.data <- read.csv("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

class(NTL.phys.data$sampledate)
# Format sampledate as date
NTL.phys.data$sampledate <- as.Date(NTL.phys.data$sampledate, format = "%m/%d/%y")
```


## Filter

Filtering allows us to choose certain rows (observations) in our dataset.

```{r}
class(NTL.phys.data$lakeid)
class(NTL.phys.data$depth)

# dplyr filtering
NTL.phys.data.surface <- filter(NTL.phys.data, depth == 0) #want data for when the depth value is 0

# Choose multiple conditions to filter
summary(NTL.phys.data$lakename) #shows all the lakes that there are
NTL.phys.data.PeterPaul <- filter(NTL.phys.data, lakename %in% c("Paul Lake", "Peter Lake")) #can have more than two names
#can be used to establish a range that we want to select
#c creates vectors with more than one element

# Choose a range of conditions of a numeric or integer variable
summary(NTL.phys.data$daynum)
NTL.phys.data.JunethruOctober <- filter(NTL.phys.data, daynum %in% c(152:304)) # everything included in the rance c() for the column daynum

# Exercise 1: 
# filter NTL.phys.data for the year 1999
# what code do you need to use, based on the class of the variable?
NTL.phys.data.year1999 <- filter(NTL.phys.data, year4 == 1999)

# Exercise 2: 
# filter NTL.phys.data for Tuesday Lake from 1990 through 1999.
NTL.phys.data.TuesdayLake <- filter(NTL.phys.data, year4 %in% c(1990:1999), lakename %in% c("Tuesday Lake"))


```
Question: Why don't we filter using row numbers?

> Answer: There is a lot of data and it would be hard to know where the data in each row ends or starts. It is much easier to filter based on what we want to pull.

## Pipes

Pipe is another method to wrangle datasets that looks cleaner and is easier to read.  We designate a pipe with `%>%`. A good way to think about the function of a pipe is with the word "then." 

Let's say we want to take our raw dataset (NTL.phys.data), *then* filter the data for Peter and Paul lakes, *then* select temperature and observation information, and *then* add a column for temperature in Fahrenheit: 

```{r}
NTL.phys.data.processed <- 
  NTL.phys.data %>%
  filter(lakename == "Paul Lake" | lakename == "Peter Lake") %>%
  select(lakename, sampledate:temperature_C) %>%
  mutate(temperature_F = (temperature_C*9/5) + 32)

#Exercise 3: Using a pipe filter NTL.phys.data for Tuesday Lake from 1990 through 1999 only for July 
NTL.phys.data.pipe.90s.july <- 
  NTL.phys.data %>%
  filter(lakename == "Tuesday Lake") %>%
  filter(year4 %in% c(1990:1999)) %>%
  mutate(sample_month = month(sampledate)) %>%
  filter(sample_month == 7)
View(NTL.phys.data.pipe.90s.july)

#or
NTL.phys.data.pipe.90s.july2 <- 
  NTL.phys.data %>%
  filter(lakename == "Tuesday Lake") %>%
  filter(year4 %in% c(1990:1999)) %>%
  filter(month(sampledat) == 7)




#Exercise 4: Using the data from part 3, a pipe and the summarise() function find the mean surface temperature (hints: you will need to add another filter for depth==0). Make sure you eliminate NAs before computing the means

NTL.phys.data.Ex4 <-
  NTL.phys.data.pipe.90s.july %>%
  filter(depth == 0) %>%
  filter(!is.na(temperature_C)) %>%
  summarise(mean=mean(temperature_C))

  
```

## Gather and Spread

For gather we will use `pivot_longer` and for spread we will use `pivot_wider`.

```{r}
#Exercise 5: gather irradiance data (measured in the water column and measured on the deck of the sampling boat) into one column using pivot_longer
NTL.phys.data.gathered <- pivot_longer(NTL.phys.data, irradianceWater:irradianeDeck, names_to= "irridance", values_to = "radiation")

#Exercise 6: spread temperatureC into more than one column based on the depth
NTL.phys.data.spread <- pivot_wider(NTL.phys.data, names_from = "depth", values_from = "temperature_C")

```

