---
title: "Assignment 5: Data Visualization"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.

##My Set Up
```{r}
#pov: me hoping this works because nothing else does
install.packages("formatR")
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE, echo=TRUE)
```


## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
#setwd
getwd()
setwd("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022")

#install packages
install.packages("tidyverse")
install.packages("cowplot")

library(tidyverse)
library(cowplot)

#load the data
peterpaullakes <- read.csv("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)
niwotridegelitter <- read.csv("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)


#2 
#both set to characters instead of date == need to change
peterpaullakes$sampledate <- as.Date(peterpaullakes$sampledate, format = "%Y-%m-%d")
niwotridegelitter$collectDate <- as.Date(niwotridegelitter$collectDate, format = "%Y-%m-%d")


```
## Define your theme
3. Build a theme and set it as your default theme.

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
#3
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
phosphoorusVSphosphate <- 
  ggplot(peterpaullakes, aes(x = po4, y = tp_ug)) +
  geom_point(aes(color = lakename)) +
  ylab("Total Phosphorus") +
  xlab("Phosphate") +
  xlim(0, 75) +
  ylim(0, 175) +
  geom_smooth(method = lm, color = "black")
print(phosphoorusVSphosphate)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5

peterpaullakes$month <- as.factor(peterpaullakes$month)


#boxplot: temp
temp <- ggplot(peterpaullakes, aes(x=month, y = temperature_C)) +
  geom_boxplot(aes(color = lakename))+
  xlab("Month") +
  ylab("Temperature (˚C)") +
  labs(color = "Lake Name")
print(temp)

#boxplot: TP
TP <- ggplot(peterpaullakes, aes(x=month, y = tp_ug)) +
  geom_boxplot(aes(color = lakename))+
  xlab("Month") +
  ylab("Total Phosphorus") +
  labs(color = "Lake Name")
print(TP)

#boxplot: TN
TN <- ggplot(peterpaullakes, aes(x=month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename))+
  xlab("Month") +
  ylab("Total Nitrogen") +
  labs(color = "Lake Name")
print(TN)

#cowplot
#save three plots as object
plot_grid()
plot_grid(NvsP2, Nutrientplot, nrow = 2, align = 'h', rel_heights = c(1.25, 1))

#create another object just for legend 
cowplot_legend <- get_legend(temp)

#use plot grid again with two rows -- one for object with three plots and one for the object with the legend
plot_grid(___, cowplot_legend, nrow=2, rel_height = c(3, 0.3)) 


```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6
questionSix <- 
  ggplot(subset(niwotridegelitter, functionalGroup == "Needles"),
         aes(x=collectDate, y=dryMass)) +
  geom_point(aes(color = nlcdClass))
print(questionSix)

#7
questionSeven <-
  ggplot(subset(niwotridegelitter, functionalGroup == "Needles"),
         aes(x=collectDate, y=dryMass)) +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(questionSeven)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
