---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Britney Pepper"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Monday, February 28 at 7:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

```{r - setting up}
#1
getwd()
#setwd("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022")


#install.packages("dplyr")
#install.packages("agricolae")
#install.packages("tinytex")
#install.packages("ggplot2")
#install.packages("tidyverse")


library(tidyverse)
library(dplyr)
library(agricolae)
library(tinytex)
library(ggplot2)
library(magrittr)
library(lubridate)

ntl_lter <- read.csv("/Users/britneypepper/Desktop/ENVIRON 872 and L/GitHubRepositories/Environmental_Data_Analytics_2022/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", 
                     stringsAsFactors = TRUE)
ntl_lter$sampledate <- as.Date(ntl_lter$sampledate, format = "%m/%d/%y")

```


2. Build a ggplot theme and set it as your default theme.

```{r - ggplot}
#2
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no difference in the mean lake temperature with depth across all lakes during the month of July.
Ha: There is a difference in the mean lake temperature with depth across all lakes during the month of July.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
ntl_lter_july <- 
  ntl_lter %>% 
  mutate(sample_month = month(sampledate)) %>%
  filter(sample_month == 7) %>%
  select(`lakename`, `year4`, `daynum`, `depth`, `temperature_C`) %>%
  na.omit()

#5
ntl_lter_july_scatter <- 
  ggplot(ntl_lter_july, aes(x = depth, y = temperature_C)) +
  ylim(0, 35) + 
  labs(color = "Temperature (C)") +
  xlab("Depth (m)") +
  ylab("Temperature (C)") +
  geom_smooth(method = "lm") +
  geom_point(aes(color = temperature_C))
print(ntl_lter_july_scatter) 
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: 
This figure suggests that there is a negative correlation between depth and temperature. As depth increases the temperature decreases. The distribution of points suggest that the trend may not be linear given that the distribution has different temperature ranges at different depths and it looks like the temperature decreases at a greater rate around 2 meters.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lr.depth.temp <- lm(data = ntl_lter_july, temperature_C ~ depth)
summary(lr.depth.temp)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 
The R-squared value is 0.7387. This means that 73.8% variability in temperature is explained by changes in depth. The degrees of freedom on which the findings are based is 9726. The residual standard error is 3.835. The p-value is less than 0.05, so the results were statistically significant. This means that we would reject the null hypothesis and say that there is a difference in the mean lake temperature with depth across all lakes during the month of July. The temperature is predicted to decrease by 1.95 degrees Celcius for every 1m change in depth.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
temp_AIC <- lm(data = ntl_lter_july, temperature_C ~ year4 + daynum + depth)
step(temp_AIC)


#10
temp_regression <- lm(data = ntl_lter_july, 
                   temperature_C ~ year4 + daynum + depth)
summary(temp_regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
AIC suggests you would not want to get rid of anything. When you run the step, the AIC increases if you remove one of the variables, so you would not want to remove anything. However, the r-squared value increased slightly to 0.7411, which means that a greater variability in temperature is explained by changes in year4, daynum, and depth.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
temp_totals <- ntl_lter_july %>%
  group_by(lakename, year4, daynum, depth) %>%
  summarise(temperature_C = sum(temperature_C))

#anova
temp_totals_anova <- aov(data = temp_totals, temperature_C ~ lakename)
summary(temp_totals_anova)

#anova as lm
temp_totals_anova_lm <- lm(data = temp_totals, temperature_C ~ lakename)
summary(temp_totals_anova_lm)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 
> Answer: There was a signifiant difference in the mean temperature among the lakes. The p-values for both tests were less than 0.05, so the mean temperature was significantly different and we would reject the null hypothesis.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
ntl_lter_lake_temp_scatter <- 
  ggplot(temp_totals, aes(x = depth, y = temperature_C, color = lakename)) +
  ylim(0, 35) + 
  ylab("Temperature (C)") + 
  xlab("Depth (m)") +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha =0.5) +
  theme(legend.position = c(0.85, 0.65))
print(ntl_lter_lake_temp_scatter) 

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(temp_totals_anova)

HSD.temp <- HSD.test(temp_totals_anova, "lakename", group = TRUE)
print(HSD.temp)


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Based on the Tukey's HSD test, the lakes that have the same mean temperature as Peter Lake is Paul Lake and Ward Lake. None of the lakes have a mean temperature that is statistically different from all the other lakes.
 
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer:
If we were just looking at Peter and Paul Lake, a test I might explore would be a two-sample t-test. This is because it would check if mean of two samples (samples from the two lakes) is equivalent.


